Parsimonious Representation of Nonlinear Dynamical Systems using Diffusion Maps: A Chemotaxis Case Study

### 2015-03-11 chemotaxis/paper/draft1.tex
transition simulations, etc.

Given the eigenvectors $\phi_1, \phi_2, \dots, \phi_{m-1} \in \mathbb{R}^m$, we would like to automatically deduce which ones capture new directions in the data, and which ones are merely repeated eigendirections. 
%
This problem was addressed previously in \cite{...} by performing successive iterations of diffusion maps, interspersed with advection along the first eigendirection at each iteration. 
%
However, both the advection procedure and the successive eigendecompositions are very expensive and intractable for larger data set. 
TODO: find other advantages of our proposed method. 
%
Here, we propose an alternative approach to address the problem of repeated eigendirections. 
%
We will attempt to fit a function $f(\phi_1, \dots, \phi_{k-1})$ to $\phi_{k}$. 
%
If the resulting fit is accurate, then we assume $\phi_{k}$ is a repeated eigendirection of $\phi_1, \dots, \phi_{k-1}$. 
%
We will use a local linear function 
\begin{equation}
\phi_k \approx \alpha + \beta^T \Phi_{k-1}
\end{equation}
%
as our functional approximation, where 
%
$\Phi_{k-1} = \begin{bmatrix} \phi_1 & \dots & \phi_{k-1} \end{bmatrix}^T$,
$\alpha \in \mathbb{R}$, and $\beta \in \mathbb{R}^{k-1}$. 
%
The coefficients $\alpha$ and $\beta$ are functions of $\Phi_{k-1}$ because we use a {\em local} linear fit, and so the coefficients change as a function of the domain.

At each point $\Phi_{k-1}(i)$, we approximate $\phi_k(i)$ by fitting a local linear function using the remaining $m-1$ data points. 
%
We solve the following optimization problem 
\begin{equation} \label{eq:opt_problem}
\hat{\alpha}_k (i) , \hat{\beta}_k(i)  = \argmin_{\alpha, \beta} \sum_{j \ne i} K(\Phi_{k-1}(i), \Phi_{k-1}(j)) \left( \phi_{k}(j) - (\alpha + \beta^T \Phi_{k-1}(j)) \right)^2.
\end{equation}
%
where $K$ is a kernel weighting function.
%
We will use a Gaussian kernel, 
%
\begin{equation}
K(\Phi_{k-1}(i), \Phi_{k-1}(j))  = \exp \left( - \frac{\|\Phi_{k-1}(i) - \Phi_{k-1} (j) \|^2}{\epsilon_{reg}^2} \right),
\end{equation}
%
where $\epsilon_{reg}$ is the kernel scale. 
%
We typically take $\epsilon_{reg} = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k-1}(i)$, as we empirically found this choice to yield good results. 
%
We then define the normalized leave-one-out cross-validation error for this local linear fit as
\begin{equation} \label{eq:cv_error}
r_{k} = \sqrt{ \frac{\sum_{i=1}^n \left( \phi_{k} (i) - (\hat{\alpha}_k(i) + \hat{\beta}_k(i)^T \Phi_{k-1}(i))  \right)^2} {\sum_{i=1}^n  \left( \phi_{k} (i) \right)^2 }}
\end{equation}
%
Note that a small value of $r_k$ implies that $\phi_{k}$ can be accurately approximated from $\phi_1, \dots, \phi_{k-1}$, and is therefore a harmonic of a previous mode.
%
Conversely, a large value of $r_{k}$ indicates that $\phi_{k}$ parameterizes a new direction in the data.
%
Because $r_{k} \approx 1$ if $\phi_{k}$ parameterizes a new direction, we choose to define $r_1 = 1$.

@@

Once the eigenvectors which parameterize unique eigendirections along the manifold have been determined, the corresponding eigenvalues can be used to calculate the relative lengths along these directions. 
%
Let $\mu_{i_1}, \mu_{i_2}, \dots$ denote the eigenvalues corresponding to eigenvectors which parameterize unique eigendirections (i.e., those eigenvectors where $r_{i_j}$ is large). 
%
From \eqref{eq:evals} and \eqref{eq:evals_relationship}, we propose to approximate the relative lengths $L_j$  along the manifold by
\begin{equation} \label{eq:est_lengths}
L_j \propto \frac{1}{\sqrt{-\log \mu_{i_j}}}.
\end{equation}
%
This measure is used to determine how many components are required to retain most of the information within the data set.
%
For example, if $L_j$ is smaller than a predefined threshold, then this component can be disregarded.

@@

\begin{figure}[t]
\centering
\begin{subfigure}{0.3\textwidth}
\includegraphics[height=2.5cm]{strip_data_L2}
\includegraphics[height=3cm]{strip_spectrum_L2}
\caption{}
\end{subfigure}
%
\hfill
%
\begin{subfigure}{0.3\textwidth}
\includegraphics[height=2.5cm]{strip_data_L4}
\includegraphics[height=3cm]{strip_spectrum_L4}
\caption{}
\end{subfigure}
%
\hfill
%
\begin{subfigure}{0.35\textwidth}
\includegraphics[height=2.5cm]{strip_data_L8}
\includegraphics[height=3cm]{strip_spectrum_L8}
\caption{}
\end{subfigure}
%
\caption{Data sets (top) and eigenvalue spectra (bottom) for strips with (a) $L_1 = 2$, $L_2 = 1$, (b) $L_1 = 4$, $L_2 = 1$, (c) $L_1 = 8$, $L_2 = 1$. The empirical eigenvalues are plotted in blue, and the analytical eigenvalues are plotted in red. From the eigenvalues which are identified as important/non-harmonics, the estimated length ratio is (a) 2.0, (b) 4.4, (c) 8.8.  TODO: normalize colors}
\label{fig:strip_compare_analytic}
\end{figure}

@@

Our model problem describes the process of cellular chemotaxis \cite{othmer2000diffusion}, where biological cells work together and exhibit coherent macroscopic dynamics regulated by extracellular sensed signals in order to accomplish tasks such as finding food or navigating away from toxins.
%
Several microscopic models have been proposed to describe chemotaxis dynamics \cite{othmer1988models, codling2008random}.
%
We will analyze one such model described by a one-dimensional velocity jump process \cite{othmer2000diffusion}.
%
%In this model, each cell is initialized with a position and a velocity on a line, and the dynamics of each cell are governed by a stochastic process.
%
%At random times, a cell will ``turn around'' and switch the direction of its velocity (this turning is controlled by extracellular signals). 
%
%We will analyze the dynamics of collections of such cells/particles. 
%
%In this example, the macroscopic dynamics of the group of cells vary depending on the value of a single system parameter.
%
%We will demonstrate how we can use diffusion maps \cite{coifman2005geometric}, a manifold learning technique, to {\em automatically} detect changes in the system's macroscopic dynamics from microscale data.
%
%This case study exhibits several key points which allow us to demonstrate the strength of our diffusion maps approach.
%
%First, this system has random microscopic behavior with smooth macroscopic dynamics governed by a single parameter which determines the regime/mode of the system. 
%
%Second, it consists of an ensemble of individual stochastic particles which must be described using the appropriate observers.
%
%Third, it has an analytic macroscopic description, which serves as a ground truth and will allow us to verify our results which are obtained in an unsupervised manner.
This specific example has an analytic macroscopic description in which the macroscopic dynamics of the group of cells vary depending on the value of a single system parameter.
%
This model serves as a ground truth and will allow us to verify our results which are obtained in an unsupervised manner.
%
Thus far, we only considered synthetic data sets for which the Euclidean distance between data points served as an informative metric. 
%
Here, we will show that our approach, when utilizing the appropriate statistical observers and affinity metric between pairs of observations, uncovers a parameterization of the microscopic data which is consistent with the macroscopic model.
%
Furthermore, we will show that changes in dynamical behavior resulting from changes in the system parameters can be automatically detected.

@@

The microscopic model consists of a collection of $N$ cells whose states are defined by their positions and velocities on a line, and the dynamics of each cell are governed by a stochastic process.
%
Let $x_i(t)$ and $v_i(t)$ denote the position and velocity, respectively, of cell $i$ at time $t$.
%
The velocity of each cell is $\pm s$, where $s$ is a (fixed) speed. 
%
We initialize the cells such that
\begin{equation}\label{eqn:system}
\begin{aligned}
x_i(0) & = 0 \\
\mathbb{P} \{ v_i(0) = +s \} & = p
\end{aligned}
\end{equation}
where $0 < p < 1$ is the probability of a cell initially moving to the right.
%
At random times, a cell will ``turn around'' and switch the direction of its velocity (this turning is controlled by extracellular signals). 
%
The velocity of each cell randomly switches between $\pm s$ following an (independent) Poisson process with rate $\lambda$.
%
We note that we have chosen a very specific one-parameter family of initial conditions which lead to simple dynamics and allow us to illustrate of our main points;
however, our analysis is not restricted to such specific cases and more complex initial conditions could be used.

@@

For the chemotaxis example, we use histograms of the particle positions as observers \cite{talmon2013empirical}. 
%
Histograms are invariant to the indexing of the particles, while retaining the spatial locations of the particles.
%
The histograms are also robust to noise \cite{talmon2013empirical}. 
%
Instead of the standard Euclidean distance, we use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between pairs of histograms. 
%
Conceptually, the EMD measures how much ``work'' it takes to transform one probability density into another.
%
It therefore not only considers where the densities are inconsistent, but also how far apart the inconsistencies are.
%
Although the brute-force computation of the EMD is computationally expensive, there has been a plethora of work in developing efficient algorithms for its approximation \cite{Pele-eccv2008, Pele-iccv2009}.
%
For the special case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions of the data \cite{rubner2000perceptual}, which can be approximated from histograms as
\begin{equation}
\| z(i) - z(j) \|_{EMD} \approx \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|
\end{equation}
where the histograms $z(i), z(j)$ are defined on equally-spaced bins in $\mathbb{R}$, and $z_k$ denotes the $k^{th}$ bin. 
%
We note that although histograms are not essential for using EMD, histograms make the computation feasible.

@@

\begin{figure}[t!]
\centering
\begin{subfigure}{5cm}
\includegraphics[width=\textwidth]{EMD_withhist_p_1}
\caption{}
\label{subfig:small_lambda_p}
\end{subfigure}
\begin{subfigure}{5cm}
\includegraphics[width=\textwidth]{EMD_withhist_t_1}
\caption{}
\label{subfig:small_lambda_t}
\end{subfigure}
\begin{subfigure}{5cm}
\includegraphics[width=\textwidth]{EMD_withhist_p_400}
\caption{}
\label{subfig:large_lambda_p}
\end{subfigure}
\begin{subfigure}{5cm}
\includegraphics[width=\textwidth]{EMD_withhist_t_400}
\caption{}
\label{subfig:large_lambda_t}
\end{subfigure}
\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with (a,b) $\lambda=1$, $s=1$, and (c,d) $\lambda=400$, $s=20$.  For each set, we run many simulations, each with $N=1000$ particles.
%
We vary the initial conditions such that $0.1 \le p  \le 0.9$, and allow each simulation to evolve for $10$ time units.
%
We discard the initial portion of each simulation trajectory ($t < 1$), which corresponds to the initial relaxation of the system.
%
The observation $z(i) \in \mathbb{R}^n$ is the histogram of $x_1(i), x_2(i), \dots, x_N(i)$ into $n = 32$ bins.
%
The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions. The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time. Representative histograms are shown for selected data points.}
\label{fig:dmaps_embed_emd}
\end{figure}

\begin{figure}
%
\def\figheight{1.4in}
%
\begin{subfigure}[t]{0.31\textwidth}
\centering
\includegraphics[height=\figheight]{chemotaxis1_evals}
\includegraphics[height=\figheight]{chemotaxis1_embed_good}
\vspace{1.2in}
%\includegraphics[height=1.5in]{chemotaxis1_embed_bad}
\caption{}
\end{subfigure}
%
\begin{subfigure}[t]{0.31\textwidth}
\centering
\includegraphics[height=\figheight]{chemotaxis2_evals}
\includegraphics[height=\figheight]{chemotaxis2_embed_good}
\includegraphics[height=\figheight]{chemotaxis2_embed_bad}
\caption{}
\end{subfigure}
%
\begin{subfigure}[t]{0.35\textwidth}
\centering
\includegraphics[height=\figheight]{chemotaxis3_evals}
\includegraphics[height=\figheight]{chemotaxis3_embed_good}
\includegraphics[height=\figheight]{chemotaxis3_embed_bad}
\caption{}
\end{subfigure}
%
\caption{Diffusion maps embedding for three different chemotaxis simulations. (a) $\lambda = 100$, $s = 10$. (b) $\lambda = 2500$, $s = 50$. (c) $\lambda = 6400$, $s = 80$. For each simulation, the eigenvalue spectrum is shown. From the spectra, we can see that the first two components are informative for (a), components 1 and 3 are informative for (b), and components 1 and 4 are informative for (c). The corresponding embeddings are shown below. For comparison, the embeddings using the first two components are also shown in (b) and (c). TODO: change (a) to larger $\lambda$ value so that the correspondences are the same across figures.}
%
\label{fig:chemotaxis_simulations_harmonics}
\end{figure}

Figure~\ref{fig:dmaps_embed_emd} shows the results of analyzing two sets of chemotaxis simulations. 
%
One set of simulations (Figures~\ref{subfig:small_lambda_p}~and~\ref{subfig:small_lambda_t}) is for a small value of $\lambda$, and the other set (Figures~\ref{subfig:large_lambda_p}~and~\ref{subfig:large_lambda_t}) is for a large value of $\lambda$. 
%
For both sets of simulations, the macroscopic variables $p$, which controls the initial distribution of the cells, and $t$, the time, are well-correlated with the eigenvectors $\phi_1$ and $\phi_2$. 
%
In both cases, $\phi_1$ is significantly more important than $\phi_2$, as indicated by the separation between the first two eigenvalues (for the small $\lambda$ case, $\mu_1 = 0.5138$, and $\mu_2 = 0.3943$, and for the large $\lambda$ case, $\mu_1 = 0.5543$, and $\mu_2 = 0.3807$).
%
The dominant coordinate $\phi_1$ is correlated with $p$ for the small $\lambda$ case (Figure~\ref{subfig:small_lambda_p}), and correlated with $t$ for the large $\lambda$ case (Figure~\ref{subfig:large_lambda_t}), indicating that the relative importance of $p$ and $t$ changes in the two simulations.
%

As illustrated in the synthetic data sets we discussed previously, the two leading eigenvectors are not guaranteed to correspond to unique eigendirections. 
%
Figure~\ref{fig:chemotaxis_simulations_harmonics} shows the results of analyzing three different simulations which span a larger range of $\lambda$ values compared to Figure~\ref{fig:dmaps_embed_emd}. 
%
We see from the eigenspectra and the cross-validation errors $r_k$ that the second unique eigendirection moves down in the spectrum as $\lambda$ increases. 
%
Plotting the data using the two eigenvectors identified as corresponding to unique eigendirections yields two-dimensional embeddings which capture the macroscopic variables $p$ and $t$ (we only show the correspondence of the two identified unique eigendirections with $p$ in the middle row of Figure~\ref{fig:chemotaxis_simulations_harmonics} and report that the correspondence between these eigendirections and $t$ is also consistent). 
%
In contrast, plotting the data using the two leading eigenvectors $\phi_1$ and $\phi_2$ produces uninformative embeddings for large values of $\lambda$, as $\phi_2$ is a repeated eigendirection which does not parameterize any new directions in the data.
%
From the top row of Figure~\ref{fig:chemotaxis_simulations_harmonics}, we can see that the eigenvalue spectrum does not exhibit any large spectral gaps.
%
However, there is a significant gap between the eigenvalues which correspond to new directions in the data, and the corresponding eigenvectors provide us with an informative two-dimensional parameterization of the data. %which is consistent with the expected macroscopic dynamics. 
%
It now becomes clear that looking at the eigenvectors {\em modulo} repeated eigendirections is essential for extracting an informative parameterization of the data, as well as characterizing the dimensionality of the data.

\subsubsection{Detecting changes in dynamical behavior}

\begin{figure}[t] 
\centering
\includegraphics[width=0.5\textwidth]{detect_change_eigenvalues}
\caption{The first two eigenvalues from our analysis as a function of $\lambda$. Each point is averaged over $10$ experiments (as in Figure~\ref{fig:dmaps_embed_emd}, one experiment consists of data collected at varying $p$ and $t$, for a fixed value of $\lambda$), with error bars denoting one standard deviation. 
The convergence of the eigenvalues at $\lambda \approx 125$ indicates switching of the dynamical ``mode''.}
\label{fig:detect_change}
\end{figure}

>>>>

Given the eigenvectors $\phi_1, \phi_2, \dots, \phi_{m-1} \in \mathbb{R}^m$, we would like to automatically deduce which ones capture new directions in the data, and which ones are merely repeated eigendirections. 
%
This problem was addressed previously in \cite{...} by performing successive iterations of diffusion maps, interspersed with advection along the first eigendirection at each iteration. 
%
However, both the advection procedure and the successive eigendecompositions are very expensive and intractable for larger data set. 
TODO: elaborate on their method and find other advantages of our proposed method (for example, theirs is ad hoc). 
%
Here, we propose an alternative approach to address the problem of repeated eigendirections. 
%
We will attempt to fit a function $f(\phi_1, \dots, \phi_{k-1})$ to $\phi_{k}$. 
%
Our approach is motivated by simple trigonometric arguments, such as $\cos2x$ can be written as a function of $\cos x$. 
%
Thus, if the resulting fit is accurate, then we assume $\phi_{k}$ is a repeated eigendirection of $\phi_1, \dots, \phi_{k-1}$. 
%
We will use a local linear function 
\begin{equation}
\phi_k \approx \alpha + \beta^T \Phi_{k-1}
\end{equation}
%
as our functional approximation, where 
%
$\Phi_{k-1} = \begin{bmatrix} \phi_1 & \dots & \phi_{k-1} \end{bmatrix}^T$,
$\alpha \in \mathbb{R}$, and $\beta \in \mathbb{R}^{k-1}$. 
%
The coefficients $\alpha$ and $\beta$ are functions of $\Phi_{k-1}$ because we use a {\em local} linear fit in the $k-1$-dimensional $\Phi_{k-1}$ space, and so the coefficients change as a function of the domain.

At each point $\Phi_{k-1}(i)$, we approximate $\phi_k(i)$ by fitting a local linear function using the remaining $m-1$ data points. 
%
We solve the following optimization problem 
\begin{equation} \label{eq:opt_problem}
\hat{\alpha}_k (i) , \hat{\beta}_k(i)  = \argmin_{\alpha, \beta} \sum_{j \ne i} K(\Phi_{k-1}(i), \Phi_{k-1}(j)) \left( \phi_{k}(j) - (\alpha + \beta^T \Phi_{k-1}(j)) \right)^2.
\end{equation}
%
where $K$ is a kernel weighting function.
%
We will use a Gaussian kernel, 
%
\begin{equation}
K(\Phi_{k-1}(i), \Phi_{k-1}(j))  = \exp \left( - \frac{\|\Phi_{k-1}(i) - \Phi_{k-1} (j) \|^2}{\epsilon_{reg}^2} \right),
\end{equation}
%
where $\epsilon_{reg}$ is the kernel scale. 
%
We typically take $\epsilon_{reg} = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k-1}(i)$, as we empirically found this choice to yield good results. 
%
We then define the normalized leave-one-out cross-validation error for this local linear fit as
\begin{equation} \label{eq:cv_error}
r_{k} = \sqrt{ \frac{\sum_{i=1}^n \left( \phi_{k} (i) - (\hat{\alpha}_k(i) + \hat{\beta}_k(i)^T \Phi_{k-1}(i))  \right)^2} {\sum_{i=1}^n  \left( \phi_{k} (i) \right)^2 }}
\end{equation}
%
Note that a small value of $r_k$ implies that $\phi_{k}$ can be accurately approximated from $\phi_1, \dots, \phi_{k-1}$. We will assume it means that $\phi_k$ is a harmonic of previous modes, i.e., a repeated eigendirction.
%
Conversely, a large value of $r_{k}$ indicates that $\phi_{k}$ parameterizes a new direction in the data.
%
We choose to set $r_1 = 1$.

@@

For the strip example, once the eigenvectors which parameterize unique eigendirections along the manifold have been determined, the corresponding eigenvalues can be used to calculate the relative lengths along these directions. 
%
For more general manifolds, we believe the eigenvalues can still provide a measure of the relative significance of the unique eigendirections.
%
Let $\mu_{i_1}, \mu_{i_2}, \dots$ denote the eigenvalues corresponding to eigenvectors which parameterize unique eigendirections (i.e., those eigenvectors where $r_{i_j}$ is large). 
%
For the strip example, from \eqref{eq:evals} and \eqref{eq:evals_relationship}, we propose to approximate the relative lengths $L_j$  along the manifold by
\begin{equation} \label{eq:est_lengths}
L_j \propto \frac{1}{\sqrt{-\log \mu_{i_j}}}.
\end{equation}
%
This measure is used to determine how many components are required to retain most of the information within the data set.
%
For example, if $L_j$ is smaller than a predefined threshold, then this component can be disregarded.

@@

\begin{figure}[t]
\centering
\begin{subfigure}{0.3\textwidth}
\includegraphics[height=2.5cm]{strip_data_L2}
\includegraphics[height=3cm]{strip_spectrum_L2}
\caption{}
\end{subfigure}
%
\hfill
%
\begin{subfigure}{0.3\textwidth}
\includegraphics[height=2.5cm]{strip_data_L4}
\includegraphics[height=3cm]{strip_spectrum_L4}
\caption{}
\end{subfigure}
%
\hfill
%
\begin{subfigure}{0.35\textwidth}
\includegraphics[height=2.5cm]{strip_data_L8}
\includegraphics[height=3cm]{strip_spectrum_L8}
\caption{}
\end{subfigure}
%
\caption{Data sets (top) and eigenvalue spectra (bottom) for strips with (a) $L_1 = 2$, $L_2 = 1$, (b) $L_1 = 4$, $L_2 = 1$, (c) $L_1 = 8$, $L_2 = 1$. The empirical eigenvalues are plotted in blue, and the analytical eigenvalues are plotted in red. From the eigenvalues which are identified as important/non-harmonics, the estimated length ratio is (a) 2.2, (b) 4.1, (c) 8.7.}
\label{fig:strip_compare_analytic}
\end{figure}

@@

Our model problem describes the process of cellular chemotaxis \cite{othmer2000diffusion}, where biological cells exhibit coherent macroscopic dynamics regulated by extracellular sensed signals in order to accomplish tasks such as finding food or navigating away from toxins.
%
Several microscopic models have been proposed to describe chemotaxis dynamics \cite{othmer1988models, codling2008random}.
%
We will analyze one such model described by a one-dimensional velocity jump process \cite{othmer2000diffusion}.
%
%In this model, each cell is initialized with a position and a velocity on a line, and the dynamics of each cell are governed by a stochastic process.
%
%At random times, a cell will ``turn around'' and switch the direction of its velocity (this turning is controlled by extracellular signals). 
%
%We will analyze the dynamics of collections of such cells/particles. 
%
%In this example, the macroscopic dynamics of the group of cells vary depending on the value of a single system parameter.
%
%We will demonstrate how we can use diffusion maps \cite{coifman2005geometric}, a manifold learning technique, to {\em automatically} detect changes in the system's macroscopic dynamics from microscale data.
%
%This case study exhibits several key points which allow us to demonstrate the strength of our diffusion maps approach.
%
%First, this system has random microscopic behavior with smooth macroscopic dynamics governed by a single parameter which determines the regime/mode of the system. 
%
%Second, it consists of an ensemble of individual stochastic particles which must be described using the appropriate observers.
%
%Third, it has an analytic macroscopic description, which serves as a ground truth and will allow us to verify our results which are obtained in an unsupervised manner.
This specific example has an analytic macroscopic description in which the macroscopic dynamics of the group of cells depend on the value of a single system parameter.
%
This model serves as a ground truth and will allow us to verify our results which are obtained in an unsupervised manner.
%
Thus far, we only considered synthetic data sets for which the Euclidean distance between data points served as an informative metric. 
%
Here, we will show that our approach, when utilizing the appropriate statistical observers and affinity metric between pairs of observations, uncovers a parameterization of the microscopic data which is consistent with the macroscopic model.
%
Furthermore, we will show that changes in dynamical behavior resulting from changes in the system parameters can be automatically detected.

@@

The microscopic model consists of a collection of $N$ cells (for our simulations, we take $N=1000$) whose states are defined by their positions and velocities on a line, and the dynamics of each cell are governed by a stochastic process.
%
Let $x_i(t)$ and $v_i(t)$ denote the position and velocity, respectively, of cell $i$ at time $t$.
%
The velocity of each cell is $\pm s$, where $s$ is a (fixed) speed. 
%
We initialize the cells such that
\begin{equation}\label{eqn:system}
\begin{aligned}
x_i(0) & = 0 \\
\mathbb{P} \{ v_i(0) = +s \} & = p
\end{aligned}
\end{equation}
where $0 < p < 1$ is the probability of a cell initially moving to the right.
%
At random times, a cell will ``turn around'' and switch the direction of its velocity (this turning is controlled by extracellular signals). 
%
The velocity of each cell randomly switches between $\pm s$ following an (independent) Poisson process with rate $\lambda$.
%
For our specific simulations, we set $s^2/\lambda=1$, which is consistent with the analysis presented in \cite{...}. 
%
We note that we have chosen a very specific one-parameter family of initial conditions which lead to simple dynamics and allow us to illustrate of our main points;
however, our analysis is not restricted to such specific cases and more complex initial conditions could be used.

@@

For the chemotaxis example, we use histograms of the particle positions as observers \cite{talmon2013empirical}. 
%
Histograms are invariant to the indexing of the particles, while retaining the spatial locations of the particles.
%
The histograms are also robust to noise \cite{talmon2013empirical}. 
%
Instead of the standard Euclidean distance, we use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between pairs of histograms. 
%
Conceptually, the EMD measures how much ``work'' it takes to transform one probability density into another.
%
It therefore not only considers where the densities are inconsistent, but also how far apart the inconsistencies are.
%
Although the brute-force computation of the EMD is computationally expensive, there has been a plethora of work in developing efficient algorithms for its approximation \cite{Pele-eccv2008, Pele-iccv2009}.
%
For the special case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions of the data \cite{rubner2000perceptual}, which can be approximated from histograms as
\begin{equation}
\| z(i) - z(j) \|_{EMD} \approx \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|
\end{equation}
where the histograms $z(i), z(j)$ are defined on equally-spaced bins in $\mathbb{R}$, and $z_k$ denotes the $k^{th}$ bin. 
%
We note that although histograms are not essential for using EMD, histograms make the computation feasible. 
%
For our specific simulations, we will take $n=32$.

@@

\begin{figure}[t!]
\def \figwidth {3.5cm}

\centering
\begin{subfigure}{\figwidth}
\includegraphics[width=\textwidth]{EMD_withhist_p_1}
\caption{}
\label{subfig:small_lambda_p}
\end{subfigure}
\begin{subfigure}{\figwidth}
\includegraphics[width=\textwidth]{EMD_withhist_t_1}
\caption{}
\label{subfig:small_lambda_t}
\end{subfigure}
\begin{subfigure}{\figwidth}
\includegraphics[width=\textwidth]{EMD_withhist_rho_1}
\caption{}
\label{subfig:small_lambda_rho}
\end{subfigure}

\begin{subfigure}{\figwidth}
\includegraphics[width=\textwidth]{EMD_withhist_p_400}
\caption{}
\label{subfig:large_lambda_p}
\end{subfigure}
\begin{subfigure}{\figwidth}
\includegraphics[width=\textwidth]{EMD_withhist_t_400}
\caption{}
\label{subfig:large_lambda_t}
\end{subfigure}
\begin{subfigure}{\figwidth}
\includegraphics[width=\textwidth]{EMD_withhist_rho_400}
\caption{}
\label{subfig:large_lambda_rho}
\end{subfigure}
\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with (a--c) $\lambda=1$, $s=1$, and (d--f) $\lambda=400$, $s=20$.  Each data set consists of $10$ simulations, with initial conditions uniformly chosen such that $0.1 \le p  \le 0.9$. We allow each simulation to evolve for $10$ time units, with $dt=1$, and use data with $t > 0$ for analysis. The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions. The data are colored by (a, d) $p$, the initial probability of a particle moving to the right, (b, e) $t$, time. Representative histograms are shown for selected data points, and (c, f) $E \rho^+ - E \rho^-$, the difference between the average position of the left- and right-moving particles.}
\label{fig:dmaps_embed_emd}
\end{figure}

\begin{figure}
%
\def\figheight{1.4in}
%
\begin{subfigure}[t]{0.31\textwidth}
\centering
\includegraphics[height=\figheight]{chemotaxis1_evals}
\includegraphics[height=\figheight]{chemotaxis1_embed_good}
\vspace{1.2in}
%\includegraphics[height=1.5in]{chemotaxis1_embed_bad}
\caption{}
\end{subfigure}
%
\begin{subfigure}[t]{0.31\textwidth}
\centering
\includegraphics[height=\figheight]{chemotaxis2_evals}
\includegraphics[height=\figheight]{chemotaxis2_embed_good}
\includegraphics[height=\figheight]{chemotaxis2_embed_bad}
\caption{}
\end{subfigure}
%
\begin{subfigure}[t]{0.35\textwidth}
\centering
\includegraphics[height=\figheight]{chemotaxis3_evals}
\includegraphics[height=\figheight]{chemotaxis3_embed_good}
\includegraphics[height=\figheight]{chemotaxis3_embed_bad}
\caption{}
\end{subfigure}
%
\caption{Diffusion maps embedding for three different chemotaxis simulation data sets. (a) $\lambda = 100$, $s = 10$. (b) $\lambda = 1600$, $s = 40$. (c) $\lambda = 6400$, $s = 80$. Each data set consists of $10$ simulations, with initial conditions uniformly chosen such that $0.1 \le p  \le 0.9$. We allow each simulation to evolve for $10$ time units, with $dt=1$, and use data with $t > 0$ for analysis. The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions. For each data set, the eigenvalue spectrum, colored by the cross-validation error, is shown. From the spectra, we can see that the first two components are informative for (a), components 1 and 3 are informative for (b), and components 1 and 4 are informative for (c). The corresponding embeddings are shown below. For comparison, the embeddings using the first two components are also shown in (b) and (c).}
%
\label{fig:chemotaxis_simulations_harmonics}
\end{figure}

Figure~\ref{fig:dmaps_embed_emd} shows the results of analyzing two sets of chemotaxis simulations. 
%
One set of simulations (Figures~\ref{subfig:small_lambda_p}--\ref{subfig:small_lambda_rho}) is for a small value of $\lambda$, and the other set (Figures~\ref{subfig:large_lambda_p}--\ref{subfig:large_lambda_rho}) is for a large value of $\lambda$. 
%
For both sets of simulations, the macroscopic variables $p$, which controls the initial distribution of the cells, and $t$, the time, are well-correlated with the eigenvectors $\phi_1$ and $\phi_2$. 
%
%In both cases, $\phi_1$ is significantly more important than $\phi_2$, as indicated by the separation between the first two eigenvalues (for the small $\lambda$ case, $\mu_1 = 0.4828$, and $\mu_2 = 0.4287$, and for the large $\lambda$ case, $\mu_1 = 0.6288$, and $\mu_2 = 0.3236$).
%
The dominant coordinate $\phi_1$ is correlated with $p$ for the small $\lambda$ case (Figure~\ref{subfig:small_lambda_p}), and correlated with $t$ for the large $\lambda$ case (Figure~\ref{subfig:large_lambda_t}), indicating that the relative importance of $p$ and $t$ changes in the two simulations.
%
TODO: should we mention fluxes here??

As illustrated in the synthetic data sets we discussed previously, the two leading eigenvectors are not guaranteed to correspond to unique eigendirections. 
%
Figure~\ref{fig:chemotaxis_simulations_harmonics} shows the results of analyzing three different simulations which span a larger range of $\lambda$ values compared to Figure~\ref{fig:dmaps_embed_emd}. 
%
We see from the eigenspectra and the cross-validation errors $r_k$ that the second unique eigendirection moves down in the spectrum as $\lambda$ increases. 
%
Plotting the data using the two eigenvectors identified as corresponding to unique eigendirections yields two-dimensional embeddings which capture the macroscopic variables $p$ and $t$ (we only show the correspondence of the two identified unique eigendirections with $p$ in the middle row of Figure~\ref{fig:chemotaxis_simulations_harmonics} and report that the correspondence between these eigendirections and $t$ is also consistent). 
%
In contrast, plotting the data using the two leading eigenvectors $\phi_1$ and $\phi_2$ produces uninformative embeddings for large values of $\lambda$, as $\phi_2$ is a repeated eigendirection which does not parameterize any new directions in the data.
%
From the top row of Figure~\ref{fig:chemotaxis_simulations_harmonics}, we can see that the eigenvalue spectrum does not exhibit any large spectral gaps.
%
However, there is a significant gap between the eigenvalues which correspond to new directions in the data (colored in blue), and the corresponding eigenvectors provide us with an informative two-dimensional parameterization of the data. %which is consistent with the expected macroscopic dynamics. 
%
It now becomes clear that looking at the eigenvectors {\em modulo} repeated eigendirections is essential for extracting an informative parameterization of the data, as well as characterizing the dimensionality of the data.
